---
title: "Assignment 04: Text Analytics using quanteda"
author: "Muhammad Al Hasib"
format: html
editor: visual
execute:
  echo: true
  warning: false
  message: false
---

## Setup: Packages

```{r}
library(quanteda)
library(quanteda.textstats)
library(quanteda.textplots)
library(quanteda.textmodels)
library(tidyverse)
library(pdftools)
library(ggplot2)
```

------------------------------------------------------------------------

## US Presidential Inaugural Speeches

### Build corpus

```{r}
corp_inaug <- corpus(data_corpus_inaugural)
summary(corp_inaug, 10)
```

### Similarities and differences over time

```{r}
# Tokenize and create DFM
toks_inaug <- tokens(corp_inaug,
                     remove_punct = TRUE,
                     remove_numbers = TRUE,
                     remove_symbols = TRUE) %>%
  tokens_remove(stopwords("en")) %>%
  tokens_wordstem()

dfm_inaug <- dfm(toks_inaug)
dfm_inaug
```

### Top features

```{r}
# Top 20 most frequent words overall
topfeatures(dfm_inaug, 20)
```

### Wordcloud of inaugural speeches

```{r}
set.seed(6323)
textplot_wordcloud(dfm_inaug,
                   max_words = 100,
                   color = c("#2980b9", "#27ae60", "#e74c3c"))
```

### Word frequency over time

```{r}
# Track key words over time
words_time <- dfm_inaug %>%
  dfm_select(pattern = c("freedom", "nation", "war", "peace", "econom*", "terror*")) %>%
  convert(to = "data.frame") %>%
  bind_cols(docvars(dfm_inaug)) %>%
  pivot_longer(cols = c("freedom", "nation", "war", "peace", "econom", "terror"),
               names_to = "word", values_to = "count") %>%
  group_by(Year, word) %>%
  summarise(count = sum(count), .groups = "drop")

ggplot(words_time, aes(x = Year, y = count, color = word)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Key Word Frequencies in Inaugural Speeches Over Time",
       x = "Year", y = "Word Count", color = "Word") +
  theme_minimal(base_size = 14)
```

### Keyness: recent vs. early presidents

```{r}
# Compare post-2000 vs pre-2000 speeches
dfm_inaug_grouped <- dfm_inaug %>%
  dfm_group(groups = ifelse(docvars(dfm_inaug, "Year") >= 2000,
                            "Post-2000", "Pre-2000"))

keyness <- textstat_keyness(dfm_inaug_grouped, target = "Post-2000")
textplot_keyness(keyness, n = 20) +
  labs(title = "Keyness: Post-2000 vs Pre-2000 Inaugural Speeches")
```

### Document similarity among presidents

```{r}
# Cosine similarity between speeches
dfm_inaug_tfidf <- dfm_tfidf(dfm_inaug)
sim_matrix <- textstat_simil(dfm_inaug_tfidf, method = "cosine")

# Show most similar speech pairs
sim_df <- as.data.frame(as.matrix(sim_matrix)) %>%
  rownames_to_column("doc1") %>%
  pivot_longer(-doc1, names_to = "doc2", values_to = "similarity") %>%
  filter(doc1 < doc2) %>%
  arrange(desc(similarity)) %>%
  head(10)

knitr::kable(sim_df, caption = "Top 10 Most Similar Inaugural Speech Pairs (Cosine Similarity)")
```

------------------------------------------------------------------------

## US Department of Defense Reports on China (1999–2025)

### Load and build corpus from PDFs

```{r}
# Set path to USDOD folder
dod_path <- "USDOD/USDOD/"

# List all PDF files
pdf_files <- list.files(dod_path, pattern = "\\.pdf$|\\.PDF$",
                        full.names = TRUE, ignore.case = TRUE)

# Extract year from filenames
years <- str_extract(basename(pdf_files), "^[0-9]{4}")

# Remove files where year could not be extracted
valid <- !is.na(years)
pdf_files <- pdf_files[valid]
years <- years[valid]

# Read PDF text
dod_texts <- map_chr(pdf_files, function(f) {
  tryCatch(
    paste(pdf_text(f), collapse = " "),
    error = function(e) ""
  )
})

# Build corpus
corp_dod <- corpus(dod_texts,
                   docnames = paste0("DOD_", years),
                   docvars = data.frame(year = as.integer(years)))

summary(corp_dod)
```

### Tokenize and build DFM

```{r}
toks_dod <- tokens(corp_dod,
                   remove_punct = TRUE,
                   remove_numbers = TRUE,
                   remove_symbols = TRUE) %>%
  tokens_remove(stopwords("en")) %>%
  tokens_remove(c("also", "will", "may", "china", "chinese",
                  "prc", "dod", "report", "military")) %>%
  tokens_wordstem()

dfm_dod <- dfm(toks_dod)
dfm_dod
```

### Top features in DOD reports

```{r}
topfeatures(dfm_dod, 20)
```

### Wordcloud of DOD reports

```{r}
set.seed(6323)
textplot_wordcloud(dfm_dod,
                   max_words = 100,
                   color = c("#2980b9", "#e74c3c", "#e67e22"))
```

### Key terms over time

```{r}
key_terms <- c("taiwan", "nuclear", "cyber", "space", "missil", "naval", "technolog")

term_trends <- dfm_dod %>%
  dfm_select(pattern = key_terms) %>%
  convert(to = "data.frame") %>%
  bind_cols(docvars(dfm_dod)) %>%
  pivot_longer(cols = all_of(key_terms),
               names_to = "term", values_to = "count") %>%
  group_by(year, term) %>%
  summarise(count = sum(count), .groups = "drop")

ggplot(term_trends, aes(x = year, y = count, color = term)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Key Term Frequencies in DOD China Reports (1999–2025)",
       x = "Year", y = "Word Count", color = "Term") +
  theme_minimal(base_size = 14)
```

------------------------------------------------------------------------

## Wordfish

Wordfish is an unsupervised text scaling model used to estimate the ideological or positional placement of documents along a single latent dimension — without requiring any reference texts or prior labels. It was developed by Slapin and Proksch (2008) and is widely used in political science to scale party manifestos, legislative speeches, and policy documents.

The model assumes that word counts follow a Poisson distribution and estimates:

-   **θ (theta):** The document position on the latent dimension (e.g., hawkish vs. dovish, early vs. late)
-   **β (beta):** The word-level discrimination parameter — how strongly each word is associated with one end of the dimension
-   **α (alpha):** Document fixed effects (document length)
-   **ψ (psi):** Word fixed effects (overall word frequency)

In the context of DOD reports, Wordfish can reveal how the rhetorical framing of China's military threat has shifted across administrations and years.

### Run Wordfish on DOD reports

```{r}
# Trim low-frequency terms for stability
dfm_dod_trim <- dfm_trim(dfm_dod, min_termfreq = 5, min_docfreq = 2)

# Run Wordfish
set.seed(6323)
wf_model <- textmodel_wordfish(dfm_dod_trim)
summary(wf_model)
```

### Visualize document positions (θ)

```{r}
# Extract document positions
wf_positions <- data.frame(
  document = docnames(dfm_dod_trim),
  year = docvars(dfm_dod_trim, "year"),
  theta = wf_model$theta,
  se = wf_model$se.theta
)

ggplot(wf_positions, aes(x = year, y = theta)) +
  geom_point(size = 3, color = "#2980b9") +
  geom_errorbar(aes(ymin = theta - 1.96 * se,
                    ymax = theta + 1.96 * se),
                width = 0.5, color = "#2980b9", alpha = 0.6) +
  geom_line(color = "#2980b9", alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(title = "Wordfish Document Positions: DOD China Reports (1999–2025)",
       subtitle = "Estimated latent position (θ) with 95% confidence intervals",
       x = "Year", y = "Estimated Position (θ)") +
  theme_minimal(base_size = 14)
```

### Word-Level discrimination (β)

```{r}
# Plot word features by position
textplot_scale1d(wf_model,
                 margin = "features",
                 highlighted = c("taiwan", "nuclear", "cyber",
                                 "space", "missil", "technolog",
                                 "naval", "aggress", "cooper")) +
  labs(title = "Wordfish: Word Discrimination Parameters (β)",
       subtitle = "Words driving positional differences across DOD reports")
```

------------------------------------------------------------------------

## Comparing positions- Wordfish and Scaling Methods

Position comparison across documents can be achieved using several scaling methods in quanteda:

### Method 1: Wordfish (Unsupervised Scaling)

Already run above. Wordfish places each DOD report on a single latent dimension based purely on word usage patterns. Documents at opposite ends of the θ scale represent the most rhetorically distinct reports — useful for tracking how threat framing has evolved without imposing any prior assumptions.

### Method 2: Wordscores (Supervised Scaling)

Wordscores requires reference texts with known positions. For DOD reports, we can assign reference scores to the earliest (1999) and most recent (2025) reports as anchors, then estimate positions of all other reports.

```{r}
# Assign reference scores: 1999 = -1 (baseline), 2025 = +1 (most recent)
ref_scores <- rep(NA, ndoc(dfm_dod_trim))
ref_scores[docvars(dfm_dod_trim, "year") == min(docvars(dfm_dod_trim, "year"))] <- -1
ref_scores[docvars(dfm_dod_trim, "year") == max(docvars(dfm_dod_trim, "year"))] <- 1

ws_model <- textmodel_wordscores(dfm_dod_trim, y = ref_scores)

# Predict scores for all documents
ws_pred <- predict(ws_model, se.fit = TRUE, newdata = dfm_dod_trim)

ws_positions <- data.frame(
  year = docvars(dfm_dod_trim, "year"),
  score = ws_pred$fit,
  se = ws_pred$se.fit
)

ggplot(ws_positions, aes(x = year, y = score)) +
  geom_point(size = 3, color = "#e74c3c") +
  geom_errorbar(aes(ymin = score - 1.96 * se,
                    ymax = score + 1.96 * se),
                width = 0.5, color = "#e74c3c", alpha = 0.6) +
  geom_line(color = "#e74c3c", alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(title = "Wordscores Document Positions: DOD China Reports",
       subtitle = "Supervised scaling anchored at 1999 (−1) and 2025 (+1)",
       x = "Year", y = "Estimated Score") +
  theme_minimal(base_size = 14)
```

### Method 3: Comparing Wordfish vs. Wordscores

```{r}
# Combine both estimates for comparison
comparison_df <- data.frame(
  year = wf_positions$year,
  Wordfish = scale(wf_positions$theta)[, 1],
  Wordscores = scale(ws_positions$score)[, 1]
) %>%
  pivot_longer(cols = c("Wordfish", "Wordscores"),
               names_to = "method", values_to = "position")

ggplot(comparison_df, aes(x = year, y = position, color = method)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Wordfish" = "#2980b9", "Wordscores" = "#e74c3c")) +
  labs(title = "Comparing Wordfish vs. Wordscores Positions",
       subtitle = "DOD China Reports (1999–2025), both scaled to z-scores",
       x = "Year", y = "Scaled Position", color = "Method") +
  theme_minimal(base_size = 14)
```

### Interpretation

The comparison plot reveals how both methods track the rhetorical evolution of DOD reports on China's military power over time. Reports from the early 2000s tend to cluster toward one end of the positional spectrum, reflecting a relatively measured tone in the post-Cold War period. Reports from the 2010s onward show a consistent shift, reflecting growing concern over China's military modernization, cyber capabilities, space programs, and assertiveness in the Taiwan Strait. Wordfish captures this shift unsupervisedly through pure word patterns, while Wordscores confirms it through supervised anchoring. The convergence of both methods strengthens confidence in the observed trend.
