---
title: "Assignment 03: ML Exercise"
author: "Muhammad Al Hasib"
format: html
editor: visual
execute:
  echo: true
  warning: false
  message: false
---

## Setup: Packages and data

```{r}
library(haven)
library(tidyverse)
```

```{r}
TEDS_2016 <- read_stata("https://github.com/datageneration/home/blob/master/DataProgramming/data/TEDS_2016.dta?raw=true")
```

------------------------------------------------------------------------

## Multiple regression model

Using `votetsai` as the dependent variable and `age`, `edu`, and `income` as predictors.

```{r}
model <- lm(votetsai ~ age + edu + income, data = TEDS_2016)
summary(model)
```

------------------------------------------------------------------------

## The `regplot` Function

Creating a function that combines `lm()`, `plot()`, and `abline()` to plot a regression fit line.

```{r}
regplot <- function(x, y) {
  fit <- lm(y ~ x)
  plot(x, y)
  abline(fit, col = "red")
}
```

------------------------------------------------------------------------

## Running `regplot` on the dependent variable

### a. Vote Choice by Age

```{r}
regplot(TEDS_2016$age, TEDS_2016$votetsai)
title(main = "Vote Choice (Tsai) by Age",
      xlab = "Age", ylab = "Vote (0 = Other, 1 = Tsai)")
```

### b. Vote Choice by Education

```{r}
regplot(TEDS_2016$edu, TEDS_2016$votetsai)
title(main = "Vote Choice (Tsai) by Education",
      xlab = "Education", ylab = "Vote (0 = Other, 1 = Tsai)")
```

### c. Vote Choice by Income

```{r}
regplot(TEDS_2016$income, TEDS_2016$votetsai)
title(main = "Vote Choice (Tsai) by Income",
      xlab = "Income", ylab = "Vote (0 = Other, 1 = Tsai)")
```

------------------------------------------------------------------------

## The problem

The problem is that the dependent variable `votetsai` is **binary** (0 = voted for other candidate, 1 = voted for Tsai Ing-wen). Linear regression assumes the outcome is **continuous**, so applying it to a binary variable produces several issues:

-   The regression line can predict values **outside the range of 0 and 1**, which are not valid probabilities
-   The relationship between predictors and a binary outcome is inherently **non-linear**
-   The assumption of **normally distributed residuals** is violated since the DV only takes two values
-   This makes the regression line a poor and misleading fit, as visible in the plots above, where data points cluster at 0 and 1 but the fitted line cuts through the middle

------------------------------------------------------------------------

## Measures to improve prediction

Since the dependent variable is binary, **logistic regression** is the appropriate model. It models the **log-odds** of the outcome and constrains all predictions between 0 and 1.

```{r}
logit_model <- glm(votetsai ~ age + edu + income,
                   data = TEDS_2016,
                   family = binomial)
summary(logit_model)
```

### Logistic regression plots

```{r}
TEDS_2016_clean <- TEDS_2016 %>%
  dplyr::select(votetsai, age, edu, income) %>%
  drop_na()
```

```{r}
ggplot(TEDS_2016_clean, aes(x = age, y = votetsai)) +
  geom_jitter(height = 0.05, alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "glm",
              method.args = list(family = binomial),
              se = TRUE,
              color = "red") +
  labs(title = "Logistic Regression: Vote Choice (Tsai) by Age",
       x = "Age", y = "Probability of Voting for Tsai") +
  theme_minimal(base_size = 16)
```

```{r}
ggplot(TEDS_2016_clean, aes(x = edu, y = votetsai)) +
  geom_jitter(height = 0.05, alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "glm",
              method.args = list(family = binomial),
              se = TRUE,
              color = "red") +
  labs(title = "Logistic Regression: Vote Choice (Tsai) by Education",
       x = "Education", y = "Probability of Voting for Tsai") +
  theme_minimal(base_size = 16)
```

```{r}
ggplot(TEDS_2016_clean, aes(x = income, y = votetsai)) +
  geom_jitter(height = 0.05, alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "glm",
              method.args = list(family = binomial),
              se = TRUE,
              color = "red") +
  labs(title = "Logistic Regression: Vote Choice (Tsai) by Income",
       x = "Income", y = "Probability of Voting for Tsai") +
  theme_minimal(base_size = 16)
```
