---
--- title: "Assignment 03: ML Exercise" author: "Muhammad Al Hasib" format: html editor: visual execute:   echo: true   warning: false   message: false ---
---

## Setup: Packages and Data

```{r} library(haven) library(tidyverse)}
```

```{r} TEDS_2016 <- read_stata("https://github.com/datageneration/home/blob/master/DataProgramming/data/TEDS_2016.dta?raw=true")}
```

------------------------------------------------------------------------

## 1. Multiple Regression Model

I am using `votetsai` as the dependent variable and `Age`, `Education`, and `Income` as predictors.

```{r} model <- lm(votetsai ~ Age + Edu + income, data = TEDS_2016) summary(model)}
```

------------------------------------------------------------------------

## 2. The `regplot` Function

Now I am creating a function that combines `lm()`, `plot()`, and `abline()` to plot a regression fit line.

```{r} regplot <- function(x, y) {   fit <- lm(y ~ x)   plot(x, y)   abline(fit, col = "red") }
```

------------------------------------------------------------------------

## 3. Running `regplot` on the Dependent Variable

### a. Vote Choice by Age

```{r} regplot(TEDS_2016$Age, TEDS_2016$votetsai) title(main = "Vote Choice (Tsai) by Age", xlab = "Age", ylab = "Vote (0 = Other, 1 = Tsai)")}
```

### b. Vote Choice by Education

```{r} regplot(TEDS_2016$Edu, TEDS_2016$votetsai) title(main = "Vote Choice (Tsai) by Education", xlab = "Education", ylab = "Vote (0 = Other, 1 = Tsai)")}
```

### c. Vote Choice by Income

```{r} regplot(TEDS_2016$income, TEDS_2016$votetsai) title(main = "Vote Choice (Tsai) by Income", xlab = "Income", ylab = "Vote (0 = Other, 1 = Tsai)")}
```

------------------------------------------------------------------------

## 4. What is the Problem? Why?

The problem is that the dependent variable `votetsai` is binary (0 = voted for other candidate, 1 = voted for Tsai Ing-wen). Linear regression assumes the outcome is continuous, so applying it to a binary variable produces several issues:

-   The regression line can predict values outside the range of 0 and 1, which are not valid probabilities
-   The relationship between predictors and a binary outcome is inherently non-linear
-   The assumption of normally distributed residuals is violated since the DV only takes two values
-   This makes the regression line a poor and misleading fit, as visible in the plots above

------------------------------------------------------------------------

## 5. What Can Be Done to Improve Prediction?

Since the dependent variable is binary, logistic regression is the appropriate model. It models the log-odds of the outcome and constrains predictions between 0 and 1.

```{r} logit_model <- glm(votetsai ~ Age + Edu + income,                    data = TEDS_2016,                    family = binomial) summary(logit_model)}
```

We can visualize the logistic fit for Age:

```{r} TEDS_2016_clean <- TEDS_2016 %>%   dplyr::select(votetsai, Age, Edu, income) %>%   drop_na()  ggplot(TEDS_2016_clean, aes(x = Age, y = votetsai)) +   geom_jitter(height = 0.05, alpha = 0.3, color = "steelblue") +   geom_smooth(method = "glm",               method.args = list(family = binomial),               se = TRUE,               color = "red") +   labs(title = "Logistic Regression: Vote Choice (Tsai) by Age",        x = "Age", y = "Probability of Voting for Tsai") +   theme_minimal(base_size = 16)}
```

```{r} ggplot(TEDS_2016_clean, aes(x = Edu, y = votetsai)) +   geom_jitter(height = 0.05, alpha = 0.3, color = "steelblue") +   geom_smooth(method = "glm",               method.args = list(family = binomial),               se = TRUE,               color = "red") +   labs(title = "Logistic Regression: Vote Choice (Tsai) by Education",        x = "Education", y = "Probability of Voting for Tsai") +   theme_minimal(base_size = 16)}
```

```{r} ggplot(TEDS_2016_clean, aes(x = income, y = votetsai)) +   geom_jitter(height = 0.05, alpha = 0.3, color = "steelblue") +   geom_smooth(method = "glm",               method.args = list(family = binomial),               se = TRUE,               color = "red") +   labs(title = "Logistic Regression: Vote Choice (Tsai) by Income",        x = "Income", y = "Probability of Voting for Tsai") +   theme_minimal(base_size = 16)}
```
