---
title: "Assignment 01: Brush Up R and Quarto"
author: "Muhammad Al Hasib"
format: html
editor: visual
execute:
  echo: true
  warning: false
  message: false
---

## Setup: Packages and Data

```{r}
library(haven)
library(tidyverse)
```

```{r}
TEDS_2016 <- haven::read_dta("https://github.com/datageneration/home/blob/master/DataProgramming/data/TEDS_2016.dta?raw=true")
```

------------------------------------------------------------------------

## Exploratory Data Analysis

### Data Overview

```{r}
dim(TEDS_2016)
names(TEDS_2016)
glimpse(TEDS_2016)
```

```{r}
summary(TEDS_2016)
```

------------------------------------------------------------------------

## Q4. What Problems Do You Encounter When Working with the Dataset?

Several issues arise when working with the TEDS 2016 dataset:

-   **Missing values:** Several variables contain a large number of `NA` values, particularly vote-related variables such as `votetsai`, `votekmt`, and income-related variables.
-   **Labelled variables:** The dataset is imported from Stata (`.dta`), so many variables come in as `labelled` class rather than standard R factor or numeric types, requiring conversion before analysis.
-   **Categorical variables stored as numeric:** Variables like `Tondu`, `Party`, `Sex`, and `Edu` are coded as integers but represent categorical categories, which can lead to incorrect analysis if treated as continuous.
-   **Skewed distributions:** Some variables such as `income` and `age` may have skewed distributions that need to be examined before modeling.

```{r}
# Check missing values across all variables
TEDS_2016 %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(),
               names_to = "variable",
               values_to = "n_missing") %>%
  filter(n_missing > 0) %>%
  arrange(desc(n_missing))
```

```{r}
# Visualize missing values
TEDS_2016 %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(),
               names_to = "variable",
               values_to = "n_missing") %>%
  filter(n_missing > 0) %>%
  ggplot(aes(x = reorder(variable, n_missing), y = n_missing)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Missing Values by Variable",
       x = NULL, y = "Count of Missing Values") +
  theme_minimal(base_size = 16)
```

------------------------------------------------------------------------

## Q5. How to Deal with Missing Values?

There are several strategies for handling missing values depending on context:

-   **Listwise deletion (`drop_na()`):** Remove all rows with any missing values. Simple but can reduce sample size significantly.
-   **Variable-wise filtering:** Remove only rows where the specific variable of interest is missing, preserving more data for other analyses.
-   **Imputation:** Replace missing values with the mean, median, or mode for numeric variables, or use multiple imputation for more rigorous analysis.
-   **Flagging:** Create a binary indicator variable to flag whether a value was missing, which can itself be informative in modeling.

For this analysis, listwise deletion is used for modeling tasks, while missing values are retained for descriptive summaries.

```{r}
# Example: drop rows missing votetsai and key predictors
teds_clean <- TEDS_2016 %>%
  dplyr::select(votetsai, Tondu, female, DPP, age, income, edu, Taiwanese, Econ_worse) %>%
  drop_na()

cat("Rows after dropping missing values:", nrow(teds_clean), "\n")
cat("Rows removed:", nrow(TEDS_2016) - nrow(teds_clean), "\n")
```

------------------------------------------------------------------------

## Q6. Exploring the Relationship Between Tondu and Other Variables

To explore the relationship between `Tondu` (unification-independence preference) and other variables, the following methods are appropriate:

-   **Cross-tabulation / frequency tables** for categorical variables (e.g., Tondu vs. female, DPP)
-   **Boxplots or violin plots** for continuous variables (e.g., Tondu vs. age, income, edu)
-   **Correlation analysis** for numeric relationships
-   **ANOVA or chi-square tests** for inferential comparisons

```{r}
# Convert Tondu to factor with labels
TEDS_2016$Tondu <- factor(TEDS_2016$Tondu,
  levels = 1:7,
  labels = c("Unification now", "Status quo, unif. in future",
             "Status quo, decide later", "Status quo forever",
             "Status quo, indep. in future", "Independence now",
             "No response"))
```

### Tondu vs. Gender (female)

```{r}
TEDS_2016 %>%
  filter(!is.na(Tondu), !is.na(female)) %>%
  mutate(female = factor(female, levels = c(0, 1), labels = c("Male", "Female"))) %>%
  ggplot(aes(x = Tondu, fill = female)) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_fill_manual(values = c("Male" = "#2980b9", "Female" = "#e74c3c")) +
  labs(title = "Tondu by Gender",
       x = "Tondu", y = "Proportion", fill = "Gender") +
  theme_minimal(base_size = 14)
```

### Tondu vs. DPP Support

```{r}
TEDS_2016 %>%
  filter(!is.na(Tondu), !is.na(DPP)) %>%
  mutate(DPP = factor(DPP, levels = c(0, 1), labels = c("Non-DPP", "DPP"))) %>%
  ggplot(aes(x = Tondu, fill = DPP)) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_fill_manual(values = c("Non-DPP" = "#2980b9", "DPP" = "#27ae60")) +
  labs(title = "Tondu by DPP Support",
       x = "Tondu", y = "Proportion", fill = "DPP") +
  theme_minimal(base_size = 14)
```

### Tondu vs. Age

```{r}
TEDS_2016 %>%
  filter(!is.na(Tondu), !is.na(age)) %>%
  ggplot(aes(x = Tondu, y = age, fill = Tondu)) +
  geom_boxplot() +
  coord_flip() +
  labs(title = "Age Distribution by Tondu",
       x = "Tondu", y = "Age") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

### Tondu vs. Education

```{r}
TEDS_2016 %>%
  filter(!is.na(Tondu), !is.na(edu)) %>%
  ggplot(aes(x = Tondu, y = as.numeric(edu), fill = Tondu)) +
  geom_boxplot() +
  coord_flip() +
  labs(title = "Education Level by Tondu",
       x = "Tondu", y = "Education") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

### Tondu vs. Income

```{r}
TEDS_2016 %>%
  filter(!is.na(Tondu), !is.na(income)) %>%
  ggplot(aes(x = Tondu, y = as.numeric(income), fill = Tondu)) +
  geom_boxplot() +
  coord_flip() +
  labs(title = "Income Level by Tondu",
       x = "Tondu", y = "Income") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

### Tondu vs. Taiwanese Identity

```{r}
TEDS_2016 %>%
  filter(!is.na(Tondu), !is.na(Taiwanese)) %>%
  mutate(Taiwanese = factor(Taiwanese, levels = c(0, 1),
                            labels = c("Non-Taiwanese", "Taiwanese"))) %>%
  ggplot(aes(x = Tondu, fill = Taiwanese)) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_fill_manual(values = c("Non-Taiwanese" = "#e74c3c", "Taiwanese" = "#27ae60")) +
  labs(title = "Tondu by Taiwanese Identity",
       x = "Tondu", y = "Proportion", fill = "Identity") +
  theme_minimal(base_size = 14)
```

### Tondu vs. Economic Pessimism (Econ_worse)

```{r}
TEDS_2016 %>%
  filter(!is.na(Tondu), !is.na(Econ_worse)) %>%
  mutate(Econ_worse = factor(Econ_worse, levels = c(0, 1),
                             labels = c("Not worse", "Economy worse"))) %>%
  ggplot(aes(x = Tondu, fill = Econ_worse)) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_fill_manual(values = c("Not worse" = "#2980b9", "Economy worse" = "#e74c3c")) +
  labs(title = "Tondu by Economic Pessimism",
       x = "Tondu", y = "Proportion", fill = "Economy") +
  theme_minimal(base_size = 14)
```

------------------------------------------------------------------------

## Q7. The votetsai Variable

The `votetsai` variable captures whether a respondent voted for DPP candidate Tsai Ing-wen (1 = Yes, 0 = No). It is a binary outcome variable, making it suitable for logistic regression rather than linear regression.

```{r}
# Distribution of votetsai
TEDS_2016 %>%
  filter(!is.na(votetsai)) %>%
  mutate(votetsai = factor(votetsai, levels = c(0, 1),
                           labels = c("Other", "Tsai"))) %>%
  ggplot(aes(x = votetsai, fill = votetsai)) +
  geom_bar() +
  scale_fill_manual(values = c("Other" = "#2980b9", "Tsai" = "#27ae60")) +
  labs(title = "Vote Choice Distribution: Tsai Ing-wen vs. Other",
       x = "Vote Choice", y = "Count") +
  theme_minimal(base_size = 16) +
  theme(legend.position = "none")
```

```{r}
# votetsai by Tondu
TEDS_2016 %>%
  filter(!is.na(votetsai), !is.na(Tondu)) %>%
  mutate(votetsai = factor(votetsai, levels = c(0, 1),
                           labels = c("Other", "Tsai"))) %>%
  ggplot(aes(x = Tondu, fill = votetsai)) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_fill_manual(values = c("Other" = "#2980b9", "Tsai" = "#27ae60")) +
  labs(title = "Vote for Tsai by Tondu Preference",
       x = "Tondu", y = "Proportion", fill = "Vote") +
  theme_minimal(base_size = 14)
```

------------------------------------------------------------------------

## Q8. Frequency Table and Barchart of Tondu

```{r}
# Frequency table
tondu_freq <- TEDS_2016 %>%
  filter(!is.na(Tondu)) %>%
  count(Tondu) %>%
  mutate(Proportion = round(n / sum(n) * 100, 1))

knitr::kable(tondu_freq,
             caption = "Frequency Table: Tondu Variable",
             col.names = c("Category", "Count", "Percentage (%)"))
```

```{r}
# Barchart of Tondu
TEDS_2016 %>%
  filter(!is.na(Tondu)) %>%
  ggplot(aes(x = fct_infreq(Tondu), fill = Tondu)) +
  geom_bar() +
  coord_flip() +
  labs(title = "Barchart: Tondu (Unification-Independence Preference)",
       x = "Tondu Category", y = "Count") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```
